---
title: "Best Colleges for Developing NBA Talent"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



```{r,warning=FALSE,message=FALSE}
library(dplyr)
library(rvest)
library(ggplot2)
library(plotly)
library(ggrepel)

```

Every year March Madness offers us a glimpse of the future stars of the NBA and we find our March Madness brackets filled with powerhouses of college basketball. Duke, Kentucky, Kansas, UNC, Villanova and so on.

These teams are often the best so they must be the best programs at developing players between their time from high school prospects to professionals NBA players, one might believe. A quick look at the top of the draft boards and we will find it littered with players from the top programs in college basketball. 

But do these teams really develop the best players or do they recruit the best players who were already on the trajectory to the NBA. Let’s look at the rate that prospects get drafted to the NBA based on their HS rankings


```{r , echo=FALSE, ,warning=FALSE, message=FALSE}
rankings <- data.frame(RSCI = numeric(0),
                       Player = character(0),
                       Pk = numeric(0),
                       College = character(0))


for(i in 2007:2020){
  
  url <- read_html(paste0("https://www.basketball-reference.com/awards/recruit_rankings_", i,".html"))
  dat <- html_table(url, header=FALSE)[[1]]
  colnames(dat) <- dat[2,]
  dat <- dat %>% select(RSCI,Player,Pk, College) %>% tail(nrow(dat)-2)
  dat <- dat %>% mutate(College = ifelse(College == "", "No College", College),
                        Player = gsub("(college)","",Player, fixed = TRUE),
                        RSCI = as.numeric(gsub("T","",RSCI)),
                        Pk = as.numeric(Pk)) %>%
    subset(RSCI != "RSCI" & RSCI != "")
  
  rankings <- rbind(rankings,dat)
}

avg.draft.pos <- rankings %>% group_by(RSCI) %>% summarise(average = mean(Pk, na.rm = TRUE))

rankings.drafted <- rankings %>% mutate(drafted = !is.na(Pk))

rankings.drafted.grouped <- rankings.drafted %>% 
  group_by(RSCI) %>% 
  summarise(average.draft.spot = mean(Pk, na.rm = TRUE), 
            number.drafted = sum(drafted), 
            total = n()) %>%
  mutate(draft.rate = number.drafted/total)


# percent drafted by college and average ranking

drafted.rate.college <- rankings.drafted %>% 
  group_by(College) %>% 
  summarise(draft.rate = sum(drafted)/n(), 
            n = n(),
            avg.hsranking = mean(RSCI))

drafted.rate.college.plot <- drafted.rate.college %>% mutate(lab = ifelse(avg.hsranking < 33 & n > 2, College,""))

p <- ggplot(drafted.rate.college.plot, aes(x=avg.hsranking, y = draft.rate, college = College)) + 
  geom_point(alpha=0.7, aes(size = n)) +
  xlab("Average High School Ranking") +
  ylab("Draft Rate") +
  ggtitle("Draft Rates of Colleges")
 
ggplotly(p)

```

So, we can see that the colleges that recruit better high school prospects, have more of them drafted into the NBA. The schools on the left are the top programs in the country such as Duke Kentucky and UNC. These schools recruit highly ranked prospects and have a high draft rate. 

A prospect’s high school ranking will affect their chances of being drafted. Can a school be credited for recruiting a good prospect and having that prospect end up in the NBA or was that prospect going to make it no matter what school he went to? To evaluate whether a school increased a player’s chance of being drafted we need to compare how often a school’s players get drafted to how often a player of the same ranking is usually drafted. 


```{r,warning=FALSE,message=FALSE}
# adjusted draft rate

rankings.drafted.grouped <- rankings.drafted.grouped %>% mutate(draft.rate)

lo <- loess(draft.rate~RSCI , rankings.drafted.grouped, span = 0.5)
adjusted.draft.rate <- predict(lo) 
adjusted.draft.rate[87:100] <- 0.118

# Graphic for adjusted draft rate
p<- ggplot(rankings.drafted.grouped, aes(x=RSCI, y = draft.rate)) + 
  geom_col() +  
  stat_smooth(span = 0.5, se=FALSE) + 
  geom_point(aes(x=1:100, y = adjusted.draft.rate)) +
  ggtitle("Draft Rates of Top 100 Ranked HS Players") +
  ylab("Draft Rate") +
  xlab("High School Ranking")

ggplotly(p)
```
The odds of a player being drafted to the NBA decrease as their ranking gets lower. Every prospect ranked 1 or 2 have been drafted into the NBA. There is some noise in the data so the actual chances to being drafted are modeled with the overlayed line. Interesting to see how tough it is to make it to the NBA. Anyone ranked outside the top 20 in high school rankings has an outside chance of being drafted. 

Using these theoretical odds of being drafted, we can calculate how much each school increased their player’s chances of being drafted. This is turned into a draft score for each school. The higher the average draft score the better.

```{r,warning=FALSE,message=FALSE}
# adjusted weighted draft score

adj.draft.joined <- rankings.drafted.grouped %>% 
  cbind(adjusted.draft.rate) %>%
  select(RSCI,adjusted.draft.rate) %>% 
  right_join(rankings.drafted,
             by = c("RSCI" = "RSCI")) %>% 
  mutate(adjusted.weighted.draft.score = drafted-adjusted.draft.rate)

# adjusted weighted draft score by college

adj.draft.score <- adj.draft.joined %>% 
  group_by(College) %>% 
  summarise(adj.draft.score.total = sum(adjusted.weighted.draft.score),
            n = n()) %>% mutate(adj.draft.score.avg = adj.draft.score.total/n)

adj.draft.score.subset <- adj.draft.score %>% subset(n>10) 

library(DT)
df <- (adj.draft.score.subset)
DT::datatable(df)

```
