---
title: "Best Colleges for Developing NBA Talent"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```



```{r,warning=FALSE,message=FALSE}
library(dplyr)
library(rvest)
library(ggplot2)
library(plotly)
library(ggrepel)

```

Every year March Madness offers us a glimpse of the future stars of the NBA and we find our March Madness brackets filled with powerhouses of college basketball. Duke, Kentucky, Kansas, UNC, Villanova and so on.

These teams are often the best so they must be the best programs at developing players between their time from high school prospects to professionals NBA players, one might believe. A quick look at the top of the draft boards and we will find it littered with players from the top programs in college basketball. 

But do these teams really develop the best players or do they recruit the best players who were already on the trajectory to the NBA. Let’s look at the rate that prospects get drafted to the NBA based on their HS rankings


```{r , echo=FALSE, ,warning=FALSE, message=FALSE}
rankings <- data.frame(RSCI = numeric(0),
                       Player = character(0),
                       Pk = numeric(0),
                       College = character(0))


for(i in 2007:2020){
  
  url <- read_html(paste0("https://www.basketball-reference.com/awards/recruit_rankings_", i,".html"))
  dat <- html_table(url, header=FALSE)[[1]]
  colnames(dat) <- dat[2,]
  dat <- dat %>% select(RSCI,Player,Pk, College) %>% tail(nrow(dat)-2)
  dat <- dat %>% mutate(College = ifelse(College == "", "No College", College),
                        Player = gsub("(college)","",Player, fixed = TRUE),
                        RSCI = as.numeric(gsub("T","",RSCI)),
                        Pk = as.numeric(Pk)) %>%
    subset(RSCI != "RSCI" & RSCI != "")
  
  rankings <- rbind(rankings,dat)
}

avg.draft.pos <- rankings %>% group_by(RSCI) %>% summarise(average = mean(Pk, na.rm = TRUE))

rankings.drafted <- rankings %>% mutate(drafted = !is.na(Pk))

rankings.drafted.grouped <- rankings.drafted %>% 
  group_by(RSCI) %>% 
  summarise(average.draft.spot = mean(Pk, na.rm = TRUE), 
            number.drafted = sum(drafted), 
            total = n()) %>%
  mutate(draft.rate = number.drafted/total)


# percent drafted by college and average ranking

drafted.rate.college <- rankings.drafted %>% 
  group_by(College) %>% 
  summarise(draft.rate = sum(drafted)/n(), 
            n = n(),
            avg.hsranking = mean(RSCI))

drafted.rate.college.plot <- drafted.rate.college %>% mutate(lab = ifelse(avg.hsranking < 33 & n > 2, College,""))

p <- ggplot(drafted.rate.college.plot, aes(x=avg.hsranking, y = draft.rate, college = College)) + 
  geom_point(alpha=0.7, aes(size = n)) +
  xlab("Average High School Ranking") +
  ylab("Draft Rate") +
  ggtitle("Draft Rates of Colleges")
 
ggplotly(p)

```

So, we can see that the colleges that recruit better high school prospects, have more of them drafted into the NBA. The schools on the left are the top programs in the country such as Duke Kentucky and UNC. These schools recruit highly ranked prospects and have a high draft rate. 

A prospect’s high school ranking will affect their chances of being drafted. Can a school be credited for recruiting a good prospect and having that prospect end up in the NBA or was that prospect going to make it no matter what school he went to? To evaluate whether a school increased a player’s chance of being drafted we need to compare how often a school’s players get drafted to how often a player of the same ranking is usually drafted. 


```{r,warning=FALSE,message=FALSE}
# adjusted draft rate

rankings.drafted.grouped <- rankings.drafted.grouped %>% mutate(draft.rate)

lo <- loess(draft.rate~RSCI , rankings.drafted.grouped, span = 0.5)
adjusted.draft.rate <- predict(lo) 
adjusted.draft.rate[87:100] <- 0.118

# Graphic for adjusted draft rate
p<- ggplot(rankings.drafted.grouped, aes(x=RSCI, y = draft.rate)) + 
  geom_col() +  
  stat_smooth(span = 0.5, se=FALSE) + 
  geom_point(aes(x=1:100, y = adjusted.draft.rate)) +
  ggtitle("Draft Rates of Top 100 Ranked HS Players") +
  ylab("Draft Rate") +
  xlab("High School Ranking")

ggplotly(p)
```
The odds of a player being drafted to the NBA decrease as their ranking gets lower. Every prospect ranked 1 or 2 have been drafted into the NBA. There is some noise in the data so the actual chances to being drafted are modeled with the overlayed line. Interesting to see how tough it is to make it to the NBA. Anyone ranked outside the top 20 in high school rankings has an outside chance of being drafted. 

Using these theoretical odds of being drafted, we can calculate how much each school increased their player’s chances of being drafted. This is turned into a draft score for each school. The higher the average draft score the better. A positive value indicates that the school increases a players chances of getting drafted relative to other top 100 players and negative values indicate the opposite. 	

```{r,warning=FALSE,message=FALSE}
# adjusted weighted draft score

adj.draft.joined <- rankings.drafted.grouped %>% 
  cbind(adjusted.draft.rate) %>%
  select(RSCI,adjusted.draft.rate) %>% 
  right_join(rankings.drafted,
             by = c("RSCI" = "RSCI")) %>% 
  mutate(adjusted.weighted.draft.score = drafted-adjusted.draft.rate)

# adjusted weighted draft score by college

adj.draft.score <- adj.draft.joined %>% 
  group_by(College) %>% 
  summarise(adj.draft.score.total = sum(adjusted.weighted.draft.score),
            n = n()) %>% mutate(adj.draft.score.avg = adj.draft.score.total/n)

adj.draft.score.subset <- adj.draft.score %>% subset(n>10) 

library(DT)
df <- (adj.draft.score.subset%>% arrange(desc(adj.draft.score.avg)) %>% select(College,adj.draft.score.avg,n))
datatable(df, caption = "Draft Score of NCAA Colleges (Min 10 top 100 players)",style = "default")

adj.draft.score.subset.ordered <- adj.draft.score.subset%>% arrange(desc(adj.draft.score.avg))

ggplot(data = adj.draft.score.subset.ordered, aes(x= reorder(College,desc(adj.draft.score.avg)), y = adj.draft.score.avg)) + geom_bar(stat="identity") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ggtitle("Draft Scores of Top NCAA Schools")+
  xlab("School")+
  ylab("Draft Score")
```

Here we can see the top schools according to the analysis.At the top we have Purdue which is significantly higher than all the other schools. A closer look at the players who have gone through the program reveals the following

```{r}

subset(rankings.drafted, College == "Purdue")

```
Purdue has recruited 11 top 100 players and 5 have been drafted to the NBA. That is quite an impressive rate considering that most of these players have been ranked pretty low and had a small chance of being drafted.

On the other end of the spectrum, we have UNLV. The team, best known for their success in the 80s and 90s, has fallen on more difficult times as their top recruits have not had much success in making it to the NBA draft. Even some of the higher ranked prospects were not drafted such as Khem Birch and Brandon McCoy both ranked 11th of their class

```{r}
subset(rankings.drafted, College == "UNLV")

```
Lastly lets take a look at the draft scores of each school plotted with the number of top 100 prospects they have recruited. 

```{r}
p <- ggplot(adj.draft.score, aes(x= n, y = adj.draft.score.avg, college = College)) + geom_point(alpha = 0.7) +
  ggtitle("Draft Score vs Number of Top 100 Prospects") +
  ylab("Draft Score") +
  xlab("Number of Top 100 Prospects")

ggplotly(p)

```
So we can see that it is very difficult for teams to consistently have an above average draft score (0 being average) as they recruit a larger number of top 100 recruits. 

One explanation is that schools that recruit a large number of top prospects don't aren't able to focus on their players and some get overlooked. Or perhaps those schools have more one and done players. Given only few months in college, the school probably has little impact on the players development. 

The more cynical interpretation is that schools really don't have any impact on the development of players as we see schools with more top 100 prospects regress to the mean. The smaller schools are able to deviate further from the mean due to small sample size.

Whatever the case it does seem like some schools have a track record of producing NBA players and others do not. When it comes to selecting a school, top prospects have many factors to consider and hopefully this analysis will help provide more clarity to their decisions. 
